---
title: "Alpha_and_Beta_diversity"
author: "Deepan"
date: "12/27/2024"
output: html_document
---

## Load libraries

```{r, message = FALSE}
# Packages I commonly use
# install.packages("stringr")
# install.packages("ggplot2")
# install.packages("reshape")
# install.packages("reshape2")
# install.packages("gplots")
# install.packages("FactoMineR")
# install.packages("factoextra")
# install.packages("geometry")

library(geometry)
library(stringr)
library(ggplot2)
library(ggpubr)
library(FactoMineR)
library(factoextra)
library(reshape)
library(reshape2)
library(gplots)
library(RColorBrewer)
library(phyloseq)
library(vegan)
library(plyr)
```

## Define color vectors

```{r}

bpart_colors <- c("darkgoldenrod", "lightskyblue3", "indianred")

indiv_colors_RPCA <- c("1" = "firebrick4", "2" = "darkolivegreen", "3" = "midnightblue", "4" = "coral1", "5" = "lightpink1", "6" = "lightskyblue3", "7" = "goldenrod1", "8" = "sienna3", "9" = "olivedrab4", "10" = "slateblue2", "11" = "burlywood4", "12" = "aquamarine4")

indiv_colors <- c("firebrick4", "darkolivegreen", "midnightblue", "coral1", "lightpink1", "lightskyblue3",  "goldenrod1", "sienna3", "olivedrab4", "slateblue2", "burlywood4", "aquamarine4")

bpart_colors_RPCA <- c("Sebaceous" = "darkgoldenrod", "Moist" = "lightskyblue3", "Dry" = "indianred")

vec_col0 <- c("indianred"  ,  "gold"      ,    "rosybrown4"  ,   "powderblue"      ,     "lightsteelblue2"     ,     "lightpink"   ,   "darkseagreen4" , "palegreen"  ,    "peachpuff3" , "khaki4"    ,     "orchid4" , "darkolivegreen", "rosybrown3"  ,   "palevioletred2" ,
              "darkorange1"    ,      "paleturquoise3" , "khaki"     ,     "dodgerblue4" ,   "sienna2"    ,    "darkgreen"   ,   "lightsalmon"  ,  "orchid2" ,   "lightsteelblue4"  ,   "lightsalmon2" ,  "lemonchiffon3"  , "honeydew"  ,   "sienna4"   ,  "darkolivegreen2"    ,    "gold1", "ivory3")

```

## plotting themes and functions

```{r}

theme_boxp <- function(){
  
  font <- "Times New Roman"
  
  theme_minimal() %+replace%
    
    theme(axis.title.y = element_text(angle = 90, size = 10), 
          axis.title.x = element_blank(),   
          axis.text.x = element_text(angle = 45, size = 10, color = "black"), 
          axis.text.y = element_text(size = 10, color = "black"), 
          
          panel.background = element_rect(fill = "white"),  
          axis.line.x = element_line(color = "black"), 
          axis.line.y = element_line(color = "black"), 
          
          legend.text = element_text(size = 10, lineheight = 0.6), 
          legend.spacing.y = unit(0.6, "mm"), 
          legend.key.height = unit(2, "mm"), 
          legend.key=element_rect(fill="white", color="black", linewidth = 0), 
          legend.title = element_blank(), 
          strip.background = element_rect(fill = NA, linetype = "solid", size = 1, colour = "black"), 
          strip.text = element_text(size = 10, margin = margin(1,1,1,1)), 
          panel.border = element_rect(fill = NA, size = 1),
          panel.spacing.y = unit(1,"line"),
    )
}

theme_pca <- function(){
  
  font <- "Times New Roman"
  
  theme_boxp() %+replace%
    
    theme(legend.position = "none", 
          axis.title.y = element_text(angle = 90, size = 12, color = "black"),
          axis.title.x = element_text(size = 12, color = "black"),
          axis.text.x = element_text(size = 10), 
          axis.text.y = element_text(size = 10),
          )
}

lm_eqn <- function(df){
    m <- lm(y ~ x, df);
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2, 
         list(a = format(unname(coef(m)[1]), digits = 2),
              b = format(unname(coef(m)[2]), digits = 2),
             r2 = format(summary(m)$r.squared, digits = 3)))
    as.character(as.expression(eq));
}

```


```{r}

df <- read.delim2("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/skin_SMGC_Zebra60perc_formatted.tsv", check.names = FALSE, row.names = 1)

tree <-  ape::read.tree('/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/SMGC_insamples_phylo_distancetree.nwk')

metadata <- read.table("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/metadata.csv",
                    sep=",", 
                    dec=".", 
                    header = TRUE,
                    stringsAsFactors = FALSE,
                    na.strings = "",
                    check.names = FALSE, 
                    comment.char = "")[-1, ]

rownames(metadata) <- metadata$Sample

metadata <- metadata[colnames(df),]

metadata[, c(-25, -26)] <- sapply(metadata[, c(-25, -26)], as.character)

df[, 1:ncol(df)] <- sapply(df[, 1:ncol(df)], as.numeric) #make all numeric inputs

df <- as.matrix(df)

otu_table_ps <- otu_table(df, taxa_are_rows = TRUE)
sample_data_ps <- sample_data(metadata)
physeq <- phyloseq(otu_table_ps, sample_data_ps)

names <- taxa_names(physeq)

# Transform counts (if necessary)
physeq <- transform_sample_counts(physeq, as.integer, taxa_are_rows = TRUE)

taxa_names(physeq) <- names

physeq <- merge_phyloseq(physeq, phy_tree(tree))

# Estimate alpha diversity
alpha <- estimate_richness(physeq, split = TRUE, measures = NULL)

alpha$Sample <- rownames(alpha)
alpha$Sample <- str_replace(alpha$Sample, "X", "")
alpha$Sample <- str_replace_all(alpha$Sample, "\\.", "-")

alpha <- merge(alpha, metadata, by = "Sample")

```

```{r}

depth <- read.delim2("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/sequencing_depth.tsv")

alpha <- merge(alpha, depth, by.x = "Sample", by.y = "sample")

#tiff("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/TIFFs/correlation_shannonvsseqdepth.tiff", units="in", width=4, height=4, res=300)
ggplot(alpha, aes(x = Shannon, y = SequencingDepth, color = Skin_Type)) +
  geom_point(size = 2) +
  scale_color_manual(values = bpart_colors) +
  geom_smooth(method = "lm", colour = "grey20", fill = "grey60") +
  stat_cor(data = alpha[, c(7,35)], aes(x = Shannon, y = SequencingDepth), inherit.aes = FALSE,p.accuracy = 0.001, r.accuracy = 0.01, method = "spearman") +
  theme_pca()
#dev.off()

cor(alpha$Shannon, alpha$SequencingDepth, method = 'spearman')
cor(alpha$Simpson, alpha$SequencingDepth, method = 'spearman')

#tiff("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/TIFFs/correlation_simpsonvsseqdepth.tiff", units="in", width=4, height=4, res=300)
ggplot(alpha, aes(x = Simpson, y = SequencingDepth)) + 
  geom_point(aes(color = Skin_Type), size = 2) + 
  scale_color_manual(values = bpart_colors) +
  geom_smooth(method = "lm", colour = "grey20", fill = "grey60") +
  stat_cor(data = alpha[, c(8,35)], aes(x = Simpson, y = SequencingDepth), inherit.aes = FALSE,p.accuracy = 0.001, r.accuracy = 0.01, method = "spearman") +
  theme_pca()
#dev.off()


```


## Box plots for diversity metrics

### All samples
```{r}

ggplot(alpha, aes(x= PMA_treated, y= Shannon)) +
  geom_boxplot(aes(fill = PMA_treated), lwd = 0.9, fatten = TRUE) +
  geom_point() +
  labs(y = "Shannon Diversity") +
  scale_fill_manual(values = c("tomato4", "moccasin")) +
  scale_y_log10(expand = expansion(mult = .1)) + stat_compare_means() + theme_boxp()

ggplot(alpha, aes(x= PMA_treated, y= Simpson)) +
  geom_boxplot(aes(fill = PMA_treated), lwd = 0.9, fatten = TRUE) +
  geom_point() +
  labs(y = "Simpson evenness") +
  scale_fill_manual(values = c("tomato4", "moccasin")) +
  scale_y_log10(expand = expansion(mult = .1)) + stat_compare_means() + theme_boxp()

```

### Skin Type
```{r}

ggplot(alpha, aes(x= PMA_treated, y= Shannon)) +
  geom_boxplot(aes(fill = Skin_Type), lwd = 0.9, fatten = TRUE) +
  geom_point() +
  labs(y = "Shannon Diversity") +
  scale_fill_manual(values = bpart_colors) +
 scale_y_log10(expand = expansion(mult = .1)) +
  facet_wrap(.~as.factor(Skin_Type)) + stat_compare_means() + theme_boxp()

ggplot(alpha, aes(x= PMA_treated, y= Simpson)) +
  geom_boxplot(aes(fill = Skin_Type), lwd = 0.9, fatten = TRUE) +
  geom_point() +
  labs(y = "Simpson evenness") +
  scale_fill_manual(values = bpart_colors) +
 scale_y_log10(expand = expansion(mult = .1)) +
  facet_wrap(.~as.factor(Skin_Type)) + stat_compare_means() + theme_boxp()

#tiff("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/TIFFs/Shannon_diversity_bodysite.tiff", units="in", width=6, height=4, res=300)
ggplot(alpha, aes(x= ordered(PMA_treated, levels = c("Raw", "PMA")), y= Shannon)) +
  geom_boxplot(aes(fill = PMA_treated), lwd = 0.9, fatten = TRUE) +
  geom_point() +
  labs(y = "Shannon Diversity") +
  scale_fill_manual(values = c("tomato4", "moccasin")) +
 scale_y_log10(expand = expansion(mult = .1)) +
  facet_wrap(.~as.factor(BodySite_ID)) + stat_compare_means(method = "wilcox", label = "p.signif",  size = 4) + theme_boxp() + 
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        strip.text = element_text(size = 12),
        panel.grid = element_blank()
        )
#dev.off
      


tiff("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/TIFFs/simpson_diversity_bodysite.tiff", units="in", width=6, height=4, res=300)
ggplot(alpha, aes(x= ordered(PMA_treated, levels = c("Raw", "PMA")), y= Simpson)) +
  geom_boxplot(aes(fill = PMA_treated), lwd = 0.9, fatten = TRUE) +
  geom_point() +
  labs(y = "Simpson evenness") +
  scale_fill_manual(values = c("tomato4", "moccasin")) +
 scale_y_log10(expand = expansion(mult = .1)) +
    facet_wrap(.~as.factor(BodySite_ID)) + stat_compare_means(method = "wilcox", label = "p.signif",  size = 4) + theme_boxp() + 
  theme(legend.position = "none",
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        strip.text = element_text(size = 12),
        panel.grid = element_blank()
        )
dev.off()

```

### PMA treated only and Raw only
```{r}

ggplot(alpha[alpha$PMA_treated == "PMA",], aes(x= as.factor(Skin_Type), y= Shannon)) +
  geom_boxplot(aes(fill = Skin_Type), lwd = 0.9, fatten = TRUE) +
  geom_point() +
  labs(y = "Shannon Diversity PMA treated only") +
  scale_fill_manual(values = bpart_colors) +
  scale_y_log10(expand = expansion(mult = .1)) + stat_compare_means() + theme_boxp()

ggplot(alpha[alpha$PMA_treated != "PMA",], aes(x= as.factor(Skin_Type), y= Shannon)) +
  geom_boxplot(aes(fill = Skin_Type), lwd = 0.9, fatten = TRUE) +
  geom_point() +
  labs(y = "Shannon Diversity Raw only") +
  scale_fill_manual(values = bpart_colors) +
  scale_y_log10(expand = expansion(mult = .1)) + stat_compare_means() + theme_boxp()


ggplot(alpha[alpha$PMA_treated == "PMA",], aes(x= as.factor(Skin_Type), y= Simpson)) +
  geom_boxplot(aes(fill = Skin_Type), lwd = 0.9, fatten = TRUE) +
  geom_point() +
  labs(y = "Simpson Evenness PMA treated only") +
  scale_fill_manual(values = bpart_colors) +
  scale_y_log10(expand = expansion(mult = .1)) + stat_compare_means() + theme_boxp()

ggplot(alpha[alpha$PMA_treated != "PMA",], aes(x= as.factor(Skin_Type), y= Simpson)) +
  geom_boxplot(aes(fill = Skin_Type), lwd = 0.9, fatten = TRUE) +
  geom_point() +
  labs(y = "Simpson Evenness Raw only") +
  scale_fill_manual(values = bpart_colors) +
  scale_y_log10(expand = expansion(mult = .1)) + stat_compare_means() + theme_boxp()


```

## By body site


```{r}
ggplot(alpha[alpha$PMA_treated == "PMA",], aes(x= ordered(BodySite_ID, levels = c("FH", "UB", "AC", "PC", "FA", "Abd")), y= Shannon)) +
  geom_boxplot(aes(fill = Skin_Type), lwd = 0.9, fatten = TRUE) +
  geom_point() +
  labs(y = "Shannon Diversity PMA treated only") +
  scale_fill_manual(values = rev(bpart_colors)) +
  scale_y_log10(expand = expansion(mult = .1)) + stat_compare_means() + theme_boxp()


# Kruskal-Wallis Test
res.kruskal_shannon <- alpha %>%
  group_by(PMA_treated) %>%
  kruskal_test(Shannon ~ Skin_Type) %>%
  mutate(label = paste("K-W =", round(statistic,2), ", p =", signif(p, 3), sep = ""),
         x = c(rep(2,2)), 
         y = c(2.8, 2.8))  # Add formatted p-value label

res.kruskal_simpson <- alpha %>%
  group_by(PMA_treated) %>%
  kruskal_test(Simpson ~ Skin_Type) %>%
  mutate(label = paste("K-W =", round(statistic,2), ", p =", signif(p, 3), sep = ""),
         x = c(rep(2,2)), 
         y = c(1.4, 1.4))  # Add formatted p-value label


# # Dunn's Test with p-value adjustment
# pwc <- alpha %>%
#   group_by(PMA_treated) %>%
#   dunn_test(Shannon ~ BodySite_ID, p.adjust.method = "fdr") %>%
#   add_y_position()  # Dynamically calculate y.position for each facet

tiff("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/TIFFs/Shannon_diversity_bodysite_comp.tiff", units="in", width=6, height=4.5, res=300)
ggplot(alpha,  aes(x= ordered(BodySite_ID, levels = c("FH", "UB", "AC", "PC", "FA", "Abd")), y= Shannon)) +
  geom_boxplot(aes(fill = Skin_Type), lwd = 0.9, fatten = TRUE) +
  geom_point() +
  facet_wrap(~ordered(PMA_treated, levels = c("Raw", "PMA")), nrow = 2) +
  labs(y = "Shannon Diversity") +
  scale_fill_manual(values = rev(bpart_colors)) +
  scale_y_log10(expand = expansion(mult = .1)) + 
  #stat_compare_means(label.y = c(0.45,0.45)) + 
   geom_text(
    data = res.kruskal_shannon, aes(label = label, x = x, y = y),  # Adjust label position
    inherit.aes = FALSE, size = 4
  ) +
  theme_boxp() + 
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        strip.text = element_text(size = 12),
        panel.grid = element_blank(), legend.position = "none")
dev.off()

tiff("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/TIFFs/Simpson_evenness_bodysite_comp.tiff", units="in", width=6, height=4.5, res=300)
ggplot(alpha,  aes(x= ordered(BodySite_ID, levels = c("FH", "UB", "AC", "PC", "FA", "Abd")), y= Simpson)) +
  geom_boxplot(aes(fill = Skin_Type), lwd = 0.9, fatten = TRUE) +
  geom_point() +
  facet_wrap(~ordered(PMA_treated, levels = c("Raw", "PMA")), nrow = 2) +
  labs(y = "Simpson Evenness") +
  scale_fill_manual(values = rev(bpart_colors)) +
  scale_y_log10(expand = expansion(mult = .1)) + 
  #stat_compare_means(label.y = c(0.45,0.45)) + 
   geom_text(
    data = res.kruskal_simpson, aes(label = label, x = x, y = y),  # Adjust label position
    inherit.aes = FALSE, size = 4
  ) +
  theme_boxp() + 
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        strip.text = element_text(size = 12),
        panel.grid = element_blank(), legend.position = "none")
dev.off()

ggplot(alpha[alpha$PMA_treated == "PMA",],  aes(x= ordered(BodySite_ID, levels = c("FH", "UB", "AC", "PC", "FA", "Abd")), y= Shannon)) +
  geom_boxplot(aes(fill = Skin_Type), lwd = 0.9, fatten = TRUE) +
  geom_point() +
  labs(y = "Simpson Evenness PMA treated only") +
  scale_fill_manual(values = rev(bpart_colors)) +
  scale_y_log10(expand = expansion(mult = .1)) + stat_compare_means() + theme_boxp()

ggplot(alpha[alpha$PMA_treated != "PMA",],  aes(x= ordered(BodySite_ID, levels = c("FH", "UB", "AC", "PC", "FA", "Abd")), y= Shannon)) +
  geom_boxplot(aes(fill = Skin_Type), lwd = 0.9, fatten = TRUE) +
  geom_point() +
  labs(y = "Simpson Evenness Raw only") +
  scale_fill_manual(values = rev(bpart_colors)) +
  scale_y_log10(expand = expansion(mult = .1)) + stat_compare_means() + theme_boxp()
```



## Beta-diversity metrics (sparsity-controlled, non-imputation using RPCA (Martino et al. 2019))

## stats functions

```{r}

run_dist_matrix_check_stats <- function(dist_matrix, ord) {
  hist(as.numeric(dist_matrix), main = "Distribution of Pairwise Distances", xlab = "Distance")
  
  if( length(unique(ord$Treatment)) > 1){
    print(adonis2(dist_matrix ~ PMA_treated, data = metadata[metadata$Sample %in% ord$Sample,], method = "robust.aitchison"))
  }
  print(adonis2(dist_matrix ~ Skin_Type, data = metadata[metadata$Sample %in% ord$Sample,], method = "robust.aitchison"))
  
  print(adonis2(dist_matrix ~ Volunteer_ID, data = metadata[metadata$Sample %in% ord$Sample,], method = "robust.aitchison"))
}

```


### All samples

```{r, fig.height=4, fig.width=4}

rpca <- read.delim("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/taxa_all_ordination.tsv.csv", header = TRUE, sep = "\t", check.names = FALSE)[,-4] #only plotting PC1 and PC2. 

rpca$Treatment <- ifelse(str_detect(rpca$Sample, "-PMA"), "PMA", "Raw")

tiff("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/TIFFs/RPCA_all_treat.tiff", units="in", width=4, height=4, res=300)
ggplot(rpca, aes(x = PC1, y = PC2, fill = Treatment)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_point(size = 4, pch = 21) +
  scale_fill_manual(values = c("tomato4", "moccasin")) +
  labs(x = "PC1 (45.94%)", y = "PC2 (43.37%)") +
  theme_pca()
dev.off()

dist <- as.dist(read.delim("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/taxa_all_distance.tsv", header = TRUE, row.names = 1, sep = "\t", check.names = FALSE))

run_dist_matrix_check_stats(dist, rpca)

tiff("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/TIFFs/RPCA_all_histogram.tiff", units="in", width=4, height=4, res=300)
hist(as.numeric(dist), main = "Distribution of Pairwise Distances", xlab = "Distance", freq = TRUE)
dev.off

```

### PMA samples only

```{r, fig.height=4, fig.width=4}

rpca_pma <- read.delim("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/taxa_pma_ordination.tsv.csv", header = TRUE, sep = "\t", check.names = FALSE) #only plotting PC1 and PC2. 

rpca_pma$Skin_Type <- ifelse(str_detect(rpca_pma$Sample, "Abd|FA"), "Dry)",
                         ifelse(str_detect(rpca_pma$Sample, "AC|PC"), "Moist", "Sebaceous"))

rpca_pma$BS <- str_split(rpca_pma$Sample, "-", simplify = TRUE)[, 2]

rpca_pma$Volunteer_ID <- str_split(rpca_pma$Sample, "-", simplify = TRUE)[,1]

tiff("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/TIFFs/RPCA_PMA_SkinType.tiff", units="in", width=4, height=4, res=300)
ggplot(rpca_pma[, -4], aes(x = PC1, y = PC2, fill = Skin_Type)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_point(size = 4, pch = 21) +
  scale_fill_manual(values = rev(bpart_colors)) +
  labs(x = "PC1 (57.54%)", y = "PC2 (39.73%)") +
  theme_pca()
dev.off()

tiff("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/TIFFs/RPCA_PMA_ind.tiff", units="in", width=4, height=4, res=300)
ggplot(rpca_pma, aes(x = PC1, y = PC2, fill = Volunteer_ID)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_point(size = 4, pch = 21) +
  scale_fill_manual(values = indiv_colors_RPCA) +
  labs(x = "PC1 (57.54%)", y = "PC2 (39.73%)") +
  theme_pca()
dev.off()

dist <- as.dist(read.delim("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/taxa_pma_distance.tsv", header = TRUE, row.names = 1, sep = "\t", check.names = FALSE))

run_dist_matrix_check_stats(dist, rpca_pma)

```

### Raw samples only

```{r, fig.height=4, fig.width=4}

rpca_nopma <- read.delim("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/taxa_no_pma_ordination.tsv.csv", header = TRUE, sep = "\t", check.names = FALSE) #only plotting PC1 and PC2. 

rpca_nopma$Skin_Type <- ifelse(str_detect(rpca_nopma$Sample, "Abd|FA"), "Dry",
                         ifelse(str_detect(rpca_nopma$Sample, "AC|PC"), "Moist", "Sebaceous"))

rpca_nopma$Volunteer_ID <- str_split(rpca_nopma$Sample, "-", simplify = TRUE)[,1]

tiff("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/TIFFs/RPCA_Raw_SkinType.tiff", units="in", width=4, height=4, res=300)
ggplot(rpca_nopma[, -4], aes(x = PC1, y = PC2, fill = Skin_Type)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_point(size = 4, pch = 21) +
  scale_fill_manual(values = rev(bpart_colors)) +
  labs(x = "PC1 (48.24%)", y = "PC2 (32.92%)") +
  theme_pca()
dev.off()

tiff("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/TIFFs/RPCA_Raw_ind.tiff", units="in", width=4, height=4, res=300)
ggplot(rpca_nopma, aes(x = PC1, y = PC2, fill = Volunteer_ID)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_point(size = 4, pch = 21) +
  scale_fill_manual(values = indiv_colors_RPCA) +
  labs(x = "PC1 (48.24%)", y = "PC2 (32.92%)") +
  theme_pca()
dev.off()

dist <- as.dist(read.delim("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/taxa_nopma_distance.tsv", header = TRUE, row.names = 1, sep = "\t", check.names = FALSE))

run_dist_matrix_check_stats(dist, rpca_nopma)

```

## Compute average distance between all samples from same individual or same skin type

```{r}

distances_nopma <- read.table("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/taxa_nopma_distance.tsv", header = TRUE, row.names = 1, sep = "\t", check.names = FALSE)
distances_pma <- read.table("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/taxa_pma_distance.tsv", header = TRUE, row.names = 1, sep = "\t", check.names = FALSE)


# Helper functions
get_Volunteer_IDidual_distances <- function(df) {
  df$Volunteer_ID <- str_split(df$Sample, "-", simplify = TRUE)[,1]
  df$Volunteer_ID2 <- str_split(df$variable, "-", simplify = TRUE)[,1]
  df <- df[df$Volunteer_ID == df$Volunteer_ID2, ]
  df$Sample <- paste(df$Sample, df$variable, sep = "_")
  df$Volunteer_ID <- as.integer(df$Volunteer_ID)
  return(df[, c("Sample", "value", "Volunteer_ID")])
}

# Process Raw samples
distances_nopma$Sample <- rownames(distances_nopma)
df <- melt(distances_nopma)

df$Sample_BS <- str_split(df$Sample, "-", simplify = TRUE)[, 2]
df$variable_BS <- str_split(df$variable, "-", simplify = TRUE)[, 2]

inter_dist <- df[ str_split(df$Sample, "-", simplify = TRUE)[, 1] != str_split(df$variable, "-", simplify = TRUE)[, 1] & df$Sample_BS == df$variable_BS, ]
inter_dist$Indiv <- str_split(inter_dist$Sample, "-", simplify = TRUE)[, 1]

inter_dist <- aggregate(inter_dist[, c(3)], by = list(inter_dist$Indiv), mean)
colnames(inter_dist)[2] <- "inter_sample_distance" 

avg_ind <- get_Volunteer_IDidual_distances(df)
intra_dist <- aggregate(avg_ind[, c(2)], by = list(avg_ind$Volunteer_ID), mean)
colnames(intra_dist)[2] <- "intra_sample_distance"

raw_distances <- merge(intra_dist, inter_dist, by = "Group.1")
raw_distances$Group.1 <- sapply(raw_distances$Group.1, as.factor)

distances_pma$Sample <- rownames(distances_pma)
df <- melt(distances_pma)

df$Sample_BS <- str_split(df$Sample, "-", simplify = TRUE)[, 2]
df$variable_BS <- str_split(df$variable, "-", simplify = TRUE)[, 2]

inter_dist <- df[ str_split(df$Sample, "-", simplify = TRUE)[, 1] != str_split(df$variable, "-", simplify = TRUE)[, 1] & df$Sample_BS == df$variable_BS, ]
inter_dist$Indiv <- str_split(inter_dist$Sample, "-", simplify = TRUE)[, 1]

inter_dist <- aggregate(inter_dist[, c(3)], by = list(inter_dist$Indiv), mean)
colnames(inter_dist)[2] <- "inter_sample_distance" 

avg_ind <- get_Volunteer_IDidual_distances(df)
intra_dist <- aggregate(avg_ind[, c(2)], by = list(avg_ind$Volunteer_ID), mean)
colnames(intra_dist)[2] <- "intra_sample_distance"

temp <- merge(intra_dist, inter_dist, by = "Group.1")

raw_distances <- merge(raw_distances, temp, by = "Group.1", suffixes = c("-Raw", "-PMA"))

raw_distances <- melt(raw_distances)
raw_distances$Treatment <- str_split(raw_distances$variable, "-", simplify = TRUE)[, 2]

tiff("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/TIFFs/INter_vs_Intra_INdivdistances.tiff", units="in", width=4, height=5, res=300)
ggplot(raw_distances, aes(x = ordered(variable, levels = c("intra_sample_distance-Raw", "intra_sample_distance-PMA", "inter_sample_distance-Raw", "inter_sample_distance-PMA")), y = value)) +
    geom_boxplot(aes(fill = Treatment), width = 0.5, lwd = 1, fatten = TRUE) +
  geom_point() +
  scale_fill_manual(values = c("Raw" = "tomato4", "PMA" = "moccasin")) +
    stat_pvalue_manual(compare_means(value ~ variable, raw_distances, p.adjust.method = "fdr"), label = "p.adj", y.position = c(2.8, 3, 3.2, 3.4, 3.6, 3.8)) + theme_boxp() + theme(legend.position = "none", axis.text.x = element_text(angle = 90), axis.title.y = element_blank(), axis.text.y = element_text(size = 12))
dev.off()

```



```{r}

# avg_SkinType <- get_skin_type_distances(df, skin_type_patterns)
# 
# # Process PMA-treated samples
# distances_pma$Sample <- rownames(distances_pma)
# df <- melt(distances_pma)
# 
# df$Sample <- str_replace_all(df$Sample, "-PMA", "")
# df$variable <- str_replace_all(df$variable, "-PMA", "")
# 
# avg_ind_pma <- get_Volunteer_IDidual_distances(df)
# avg_SkinType_pma <- get_skin_type_distances(df, skin_type_patterns)
# 
# # Merge Raw and PMA distances
# avg_ind <- merge(avg_ind, avg_ind_pma, by = "Sample", suffixes = c("_Raw", "_PMA"))
# colnames(avg_ind) <-  str_replace(colnames(avg_ind), "value_", "")
# 
# avg_SkinType <- merge(avg_SkinType, avg_SkinType_pma, by = "Sample", suffixes = c("_Raw", "_PMA"))
# colnames(avg_SkinType) <-  str_replace(colnames(avg_SkinType), "value_", "")
# 
# 
# # Statistical Analysis
# compute_wilcox <- function(data, group_col, value1_col, value2_col) {
#   results <- data.frame()
#   grouped_data <- data %>% group_by(!!sym(group_col))
#   
#   for (group_name in unique(grouped_data[[group_col]])) {
#     group_subset <- grouped_data %>% filter(!!sym(group_col) == group_name)
#     if (nrow(group_subset) > 3) {  # Skip if insufficient data
#       wilcox_result <- wilcox.test(group_subset[[value1_col]], group_subset[[value2_col]], paired = FALSE)
#       result_row <- data.frame(
#         Group = group_name,
#         N = nrow(group_subset),
#         PValue = p.adjust( wilcox_result$p.value, method = "fdr", n = nrow(group_subset)),
#         statistic = wilcox_result$statistic
#       )
#       results <- rbind(results, result_row)
#     }
#   }
#   
#   fisher_combined_p <- -2 * sum(log(results$PValue))
#   df_combined <- data.frame(
#   statistic = sum(results$statistic),
#   combined_p = pchisq(fisher_combined_p, df = 2 * nrow(results), lower.tail = FALSE ),
#   group1 = "Raw", 
#   group2 = "PMA"
#   )
# 
#   df_combined$p.adj <- sapply(df_combined$combined_p, function(x) {
#    if (x <= 0.05) ifelse(x <= 0.01, ifelse(x <= 0.001, "***", "**"), "*") else "ns"
#    })
#   return(df_combined)
# }
# 
# 
# 
# # Plotting
# plot_distances <- function(df, results, title, y_col = "value") {
#   ggplot(melt(df), aes(x = variable, y = !!sym(y_col))) +
#     geom_boxplot(aes(fill = variable), width = 0.5, lwd = 1, fatten = TRUE) +
#     stat_pvalue_manual(results, label = "p.adj", y.position = 5) +
#     scale_fill_manual(values = c("Raw" = "tomato4", "PMA" = "moccasin")) +
#     labs(title = title, x = "", y = "") +
#     theme_pca() + theme(plot.title = element_text(size = 10))
# }
# 
# results_Volunteer_ID <- compute_wilcox(data = avg_ind, group_col = "Volunteer_ID_Raw", value1_col= "Raw", value2_col = "PMA")
# results_SkinType <- compute_wilcox(data = avg_SkinType, group_col = "Skin_Type_Raw", value1_col= "Raw", value2_col = "PMA")
# 
# 
# avg_ind[, c(-2, -4)] <- sapply(avg_ind[, c(-2, -4)], as.character)
# 
# tiff("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/TIFFs/Distance_Ind.tiff", units="in", width=4, height=4, res=300)
# plot_distances(avg_ind, results_Volunteer_ID, "Distances Between Samples from Same Volunteer_ID")
# dev.off()
# 
# tiff("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/TIFFs/Distance_SkinType.tiff", units="in", width=4, height=4, res=300)
# plot_distances(avg_SkinType, results_SkinType, "Distances Between Samples from Same Skin Type")
# dev.off()
# 
# avg_ind <- aggregate(avg_ind[, c(2, 4)], by = list(avg_ind$Volunteer_ID_Raw), mean)
# 
# avg_SkinType <- aggregate(avg_SkinType[, c(2, 4)], by = list(avg_SkinType$Skin_Type_Raw), mean)
# 
# df <- data.frame(x = c("avg_intra_indiv-raw", "avg_inter_indiv-raw", "avg_intra_indiv-PMA", "avg_inter_indiv-PMA"),
#                  y = c(mean(avg_ind$Raw, na.rm = TRUE), mean(avg_SkinType$Raw, na.rm = TRUE), mean(avg_ind$PMA, na.rm = TRUE), mean(avg_SkinType$PMA, na.rm = TRUE))
#                  )
# ggplot(df, aes(x = x, y = y)) + geom_bar(stat = "identity") + stat_compare_means(method = "wilcox.test", comparisons = list(c("avg_intra_indiv-raw", "avg_intra_indiv-PMA"), c("avg_inter_indiv-raw", "avg_inter_indiv-PMA")), label = "p.signif")
```


## Compute convex hull volume per participant from robust aitchison distance matrix for Raw vs PMA treated samples. 

```{r, fig.height=4, fig.width=4}


# Group data by your variable of interest (e.g., Volunteer or Skin Type)
compute_hull_volumes <- function(data, group_column, coord_columns) {
  group_volumes <- lapply(unique(data[[group_column]]), function(group) {
    # Subset data for the group
    group_data <- data[data[[group_column]] == group, coord_columns]
    
    # Check if there are enough points for a convex hull (k+1 points for k dimensions)
    if (nrow(group_data) < ncol(group_data) + 1) {
      return(data.frame(Group = group, Volume = NA))
    }
    
    # Compute convex hull volume
    volume <- tryCatch({
      convhulln(as.matrix(group_data), options = "FA")$vol
    }, error = function(e) {
      NA  # Return NA if convex hull cannot be computed
    })
    
    return(data.frame(Group = group, Volume = volume))
  })
  
  # Combine results into a single data frame
  do.call(rbind, group_volumes)
}

coord_columns <- c("PC1", "PC2", "PC3") 

# Compute convex hull volumes
volunteer_hullvolumes <- compute_hull_volumes(rpca_nopma[, -5], "Volunteer_ID", coord_columns)
volunteer_hullvolumes$PMA <- compute_hull_volumes(rpca_pma[, -5], "Volunteer_ID", coord_columns)$Volume

colnames(volunteer_hullvolumes) <- c("Volunteer_ID", "Raw", "PMA")
volunteer_hullvolumes <- volunteer_hullvolumes[complete.cases(volunteer_hullvolumes), ]

#normality test
hist(melt(volunteer_hullvolumes)$value, main = "Histogram", xlab = "Values")
shapiro.test(melt(volunteer_hullvolumes)$value) # not normally distributed

#tiff("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/TIFFs/Volunteer_convexhullvolume.tiff", units="in", width=4, height=4, res=300)
ggplot(melt(volunteer_hullvolumes), aes(x = variable, y = value)) + 
    geom_boxplot(aes(fill = variable), width = 0.5, lwd = 1, fatten = TRUE) +
    geom_point(aes(color = Volunteer_ID)) +
  geom_line(aes(group = Volunteer_ID), color = "grey70", size = 0.3) +
    scale_color_manual(values = indiv_colors) +
    stat_compare_means(compare_means(value~variable, melt(volunteer_hullvolumes), method = "wilcox.test", p.adjust.method = "fdr", paired = TRUE), comparisons = c("group1", "group2"), label = "p.format") +
    scale_fill_manual(values = c("Raw" = "tomato4", "PMA" = "moccasin")) +
  labs(title = "Convex-hull volume for samples from same Volunteer", y = "Robust Aitchison convex-hull volume", x = "") +
    #scale_y_log10() +
    theme_pca() +
    theme(panel.grid = element_blank(), plot.title = element_text(size = 10), axis.title.y = element_text(size = 10))
#dev.off()
wilcox.test(value ~ variable, reshape2::melt(volunteer_hullvolumes), paired = TRUE, method = 'wilcox.test')

```

## Beta-diversity metrics (non-rpca)

```{r}

dist_methods <- unlist(distanceMethodList)
dist_methods <- dist_methods[-(3)]
dist_methods = dist_methods[-which(dist_methods=="ANY")]
dist_methods <- dist_methods[(1:11)]
dist_methods <- dist_methods[!dist_methods %in% "morisita" ]

dist_methods <- dist_methods[c(7, 9)]

```

### All samples and PMA treatment effect

```{r}
Exp <- data.frame(F =1, "Pr(>F)"=2, check.names = FALSE)

sub <- physeq

plist <- vector("list", length(dist_methods))
names(plist) = dist_methods
for( i in dist_methods ){
  i
    # Calculate distance matrix
    iDist <- distance(sub, method=i)
    # Calculate ordination
    iMDS  <- ordinate(sub, "MDS", distance=iDist)
    ## Make plot
    # Don't carry over previous plot (if error, p will be blank)
    p <- NULL
    # Create plot, store as temp variable, p
    p <- plot_ordination(sub, iMDS, color="PMA_treated", shape="Skin_Type")
    # Add title to each plot
    p <- p + ggtitle(paste("MDS using distance method ", i, sep=""))
    # Save the graphic to file.
    plist[[i]] = p
    Exp <- rbind(Exp,  as.data.frame(adonis2(iDist ~ (sample_data(sub)$PMA_treated))[1, c(4,5)]))
}

Exp <- Exp[-1, ]

df = ldply(plist, function(x) x$data)
names(df)[1] <- "distance"
df$distance <- ordered(df$distance, levels =unique(df$distance))
p = ggplot(df, aes(Axis.1, Axis.2, color=PMA_treated, shape=Skin_Type)) + scale_color_manual("PMA_treated", values = c("tomato4", "moccasin"))
p = p + geom_point(size=3)
p = p + facet_wrap(~distance, scales="free", nrow = 2)
p = p + ggtitle("MDS on various distance metrics for Enterotype dataset")


rownames(Exp) <- dist_methods
colnames(Exp) <- paste("Experiment_", colnames(Exp))

#tiff("/media/deepan/Deepan/skin_ ___", units="in", width=16, height=8, res=300)
p + theme_bw() + theme(axis.text = element_text(size = 10), axis.title = element_text(size = 12),
         legend.text = element_text(size = 12), legend.title = element_text(size = 16), strip.text = element_text(size = 14),
         plot.title = element_blank())

Exp
#dev.off

```
### PMA treated samples only and clustering by SkinType and Volunteer_IDidual

```{r}

SkinType <- data.frame(F =1, "Pr(>F)"=2, check.names = FALSE)
Ind <- SkinType

sub <- subset_samples(physeq, PMA_treated=="PMA")

plist <- vector("list", length(dist_methods))
names(plist) = dist_methods
for( i in dist_methods ){
  i
    # Calculate distance matrix
    iDist <- distance(sub, method=i)
    # Calculate ordination
    iMDS  <- ordinate(sub, "MDS", distance=iDist)
    ## Make plot
    # Don't carry over previous plot (if error, p will be blank)
    p <- NULL
    # Create plot, store as temp variable, p
    p <- plot_ordination(sub, iMDS, color="Volunteer_ID", shape="Skin_Type")
    # Add title to each plot
    p <- p + ggtitle(paste("MDS using distance method ", i, sep=""))
    # Save the graphic to file.
    plist[[i]] = p
    SkinType <- rbind(SkinType,  as.data.frame(adonis2(iDist ~ (sample_data(sub)$Skin_Type))[1, c(4,5)]))
    Ind <- rbind(Ind,  as.data.frame(adonis2(iDist ~ (sample_data(sub)$Volunteer_ID))[1, c(4,5)]))
    #Med <- rbind(Med,  as.data.frame(adonis2(iDist ~ (sample_data(physeq)$Media))[1, c(4,5)]))
}

SkinType <- SkinType[-1, ]
Ind <- Ind[-1, ]


df = ldply(plist, function(x) x$data)
names(df)[1] <- "distance"
df$distance <- ordered(df$distance, levels =unique(df$distance))
p = ggplot(df, aes(Axis.1, Axis.2, color= ordered(Volunteer_ID, levels = c(1:12)), shape=Skin_Type)) + scale_color_manual("Volunteer_ID", values = indiv_colors_RPCA)
p = p + geom_point(size=3)
p = p + facet_wrap(~distance, scales="free", nrow = 2)
p = p + ggtitle("MDS on various distance metrics for Enterotype dataset")


rownames(SkinType) <- dist_methods
colnames(SkinType) <- paste("SkinType_", colnames(SkinType))

rownames(Ind) <- dist_methods
colnames(Ind) <- paste("Volunteer_IDidual_", colnames(Ind))

tiff("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/TIFFs/beta_diversity_brayjaccard_PMA.tiff", units="in", width=5, height=5, res=300)
p + theme_bw() + theme(axis.text = element_text(size = 10), axis.title = element_text(size = 12),
         legend.text = element_text(size = 12), legend.title = element_text(size = 16), strip.text = element_text(size = 14),
         plot.title = element_blank())
dev.off

cbind(SkinType, Ind)

# write.csv(cbind(Exp, Med, Cond), "/media/deepan/Deepan/skin_comm/PERMANOVA_all_beta_distances.csv", sep = ",", row.names = TRUE, col.names = TRUE)

```

### Raw samples only and clustering by SkinType and Volunteer_IDidual

```{r}

SkinType <- data.frame(F =1, "Pr(>F)"=2, check.names = FALSE)
Ind <- SkinType

sub <- subset_samples(physeq, PMA_treated=="Raw")

metadata

plist <- vector("list", length(dist_methods))
names(plist) = dist_methods
for( i in dist_methods ){
  i
    # Calculate distance matrix
    iDist <- distance(sub, method=i)
    # Calculate ordination
    iMDS  <- ordinate(sub, "MDS", distance=iDist)
    ## Make plot
    # Don't carry over previous plot (if error, p will be blank)
    p <- NULL
    # Create plot, store as temp variable, p
    p <- plot_ordination(sub, iMDS, color="Volunteer_ID", shape="Skin_Type")
    # Add title to each plot
    p <- p + ggtitle(paste("MDS using distance method ", i, sep=""))
    # Save the graphic to file.
    plist[[i]] = p
    SkinType <- rbind(SkinType,  as.data.frame(adonis2(iDist ~ (sample_data(sub)$Skin_Type))[1, c(4,5)]))
    Ind <- rbind(Ind,  as.data.frame(adonis2(iDist ~ (sample_data(sub)$Volunteer_ID))[1, c(4,5)]))
    #Med <- rbind(Med,  as.data.frame(adonis2(iDist ~ (sample_data(physeq)$Media))[1, c(4,5)]))
}

SkinType <- SkinType[-1, ]
Ind <- Ind[-1, ]


df = ldply(plist, function(x) x$data)
names(df)[1] <- "distance"
df$distance <- ordered(df$distance, levels =unique(df$distance))
p = ggplot(df, aes(Axis.1, Axis.2, color=ordered(Volunteer_ID, levels = c(1:12)), shape=Skin_Type)) + scale_color_manual("Volunteer_ID", values = indiv_colors_RPCA)
p = p + geom_point(size=3)
p = p + facet_wrap(~distance, scales="free", nrow = 2)
p = p + ggtitle("MDS on various distance metrics for Enterotype dataset")


rownames(SkinType) <- dist_methods
colnames(SkinType) <- paste("SkinType_", colnames(SkinType))

rownames(Ind) <- dist_methods
colnames(Ind) <- paste("Ind", colnames(Ind))

tiff("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/TIFFs/beta_diversity_brayjaccard_raw.tiff", units="in", width=5, height=5, res=300)
p + theme_bw() + theme(axis.text = element_text(size = 10), axis.title = element_text(size = 12), axis.text.x = element_text(angle = 30),
         legend.text = element_text(size = 12), legend.title = element_text(size = 16), strip.text = element_text(size = 14),
         plot.title = element_blank())
dev.off

cbind(SkinType, Ind)

# write.csv(cbind(Exp, Med, Cond), "/media/deepan/Deepan/skin_comm/PERMANOVA_all_beta_distances.csv", sep = ",", row.names = TRUE, col.names = TRUE)

```
