---
title: "SMGC_biomfile_preprocessing"
author: "Deepan"
date: "12/18/2024"
output: html_document
---
## Load libraries

```{r, message = FALSE}
# Packages I commonly use
# install.packages("stringr")
# install.packages("ggplot2")
# install.packages("reshape")
# install.packages("reshape2")
# install.packages("gplots")
# install.packages("FactoMineR")
# install.packages("factoextra")
library(stringr)
library(ggplot2)
library(FactoMineR)
library(factoextra)
library(reshape)
library(reshape2)
library(gplots)
library(RColorBrewer)
```

## Load the tsv file from woltka classify and make numeric

```{r}

df <- read.delim2("/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/skin_SMGC_Zebra60perc.tsv", skip = 1, check.names = FALSE)

df[, -1] <- sapply(df[, -1], as.numeric)

```

## Change the column names

```{r}

colnames(df) <- str_replace_all(colnames(df), "_PMA", "-PMA")
colnames(df) <- str_split(colnames(df), "_S", simplify = TRUE)[,1]

colnames(df) <- str_replace(colnames(df), "1bis-", "12-")

```

## Extract only the columns of 1 and 1bis for averaging

```{r}
# 
# df_one <- df[ , str_detect( colnames(df), "#OTU|1-|1bis-")]
# df_one <- df_one[,  !str_detect( colnames(df_one), "11-")]
# 
# #remove these columns from df
# 
# df <- df[ , colnames(df) %in% c("#OTU ID", colnames(df)[!colnames(df)%in%colnames(df_one)]) ]

```

## Function to keep only unique body-site volunteer 1 combos, or else calculate average [NOT DOING THIS ANYMORE]

```{r}

# Function to process the dataframe
process_dataframe <- function(df) {
  library(dplyr)

  # Extract column names
  col_names <- colnames(df)

  # Parse patient numbers and body sites
  parsed <- strcapture("^(\\d+)(bis)?-(.+)$", col_names, data.frame(Patient = character(), Bis = character(), Site = character()))
  parsed$Bis[is.na(parsed$Bis)] <- ""  # Handle NAs for easier comparisons

  # Group columns by patient and site
  result <- df %>%
    select(all_of(col_names)) %>%
    {
      # Loop through each unique combination of Patient and Site
      unique_combinations <- unique(parsed[, c("Patient", "Site")])
      purrr::map_dfc(1:nrow(unique_combinations), function(i) {
        patient <- unique_combinations$Patient[i]
        site <- unique_combinations$Site[i]

        # Find relevant columns for this combination
        cols_1 <- col_names[parsed$Patient == patient & parsed$Site == site & parsed$Bis == ""]
        cols_bis <- col_names[parsed$Patient == patient & parsed$Site == site & parsed$Bis == "bis"]

        if (length(cols_1) > 0 && length(cols_bis) > 0) {
          # Case: Both 1 and 1bis exist -> Replace with average
          colname <- paste0(patient, "-", site)
          tibble(!!colname := rowMeans(select(., all_of(c(cols_1, cols_bis)))))
        } else if (length(cols_1) == 0 && length(cols_bis) > 0) {
          # Case: Only 1bis exists -> Rename to 1
          colname <- paste0(patient, "-", site)
          rename(select(., all_of(cols_bis)), !!colname := everything())
        } else if (length(cols_1) > 0 && length(cols_bis) == 0) {
          # Case: Only 1 exists -> Keep as is
          select(., all_of(cols_1))
        }
      })
    }

  result
}

# # Process the dataframe
# processed_df <- cbind(process_dataframe(df_one[,-1]), df_one[,1])
# processed_df <- processed_df[, c(ncol(processed_df), 1:(ncol(processed_df)-1))]
# 
# df_one <-processed_df
# colnames(df_one)[1] <- "#OTU ID"

```

## Merging the processed df_one to make the correct, final tsv with all volunteers

```{r}

# df <- merge(df, df_one, by = "#OTU ID")

```

## Making the correct order of the dataframe

```{r}

colnames(df)

col_order <- c( "#OTU ID", "1-FH",  "1-FH-PMA", "1-UB", "1-UB-PMA", "1-AC",  "1-AC-PMA", "1-PC-PMA", "1-FA-PMA", "1-Abd", "1-Abd-PMA", "2-FH", "2-FH-PMA", "2-UB", "2-UB-PMA", "2-AC", "2-AC-PMA", "2-PC", "2-PC-PMA", "2-FA", "2-FA-PMA","2-Abd", "2-Abd-PMA", "3-FH", "3-FH-PMA", "3-UB", "3-UB-PMA", "3-AC", "3-AC-PMA", "3-PC", "3-PC-PMA", "3-FA", "3-FA-PMA", "3-Abd", "3-Abd-PMA", "4-FH", "4-FH-PMA", "4-UB", "4-UB-PMA", "4-AC", "4-AC-PMA", "4-PC", "4-PC-PMA", "4-FA", "4-FA-PMA", "4-Abd", "4-Abd-PMA", "5-FH", "5-FH-PMA", "5-UB", "5-UB-PMA", "5-AC", "5-AC-PMA", "5-PC", "5-PC-PMA", "5-FA", "5-FA-PMA", "5-Abd", "5-Abd-PMA", "6-FH", "6-FH-PMA", "6-UB" ,"6-UB-PMA", "6-AC", "6-AC-PMA", "6-PC", "6-PC-PMA", "6-FA", "6-FA-PMA", "6-Abd", "6-Abd-PMA", "7-FH", "7-FH-PMA", "7-UB", "7-UB-PMA", "7-AC", "7-AC-PMA", "7-PC", "7-PC-PMA", "7-FA", "7-FA-PMA", "7-Abd", "7-Abd-PMA", "8-FH", "8-FH-PMA", "8-UB", "8-UB-PMA", "8-AC", "8-AC-PMA", "8-PC", "8-PC-PMA", "8-FA", "8-FA-PMA", "8-Abd", "8-Abd-PMA", "9-FH", "9-FH-PMA", "9-UB", "9-UB-PMA", "9-AC", "9-AC-PMA", "9-PC", "9-PC-PMA", "9-FA", "9-FA-PMA", "9-Abd", "9-Abd-PMA", "10-FH", "10-FH-PMA", "10-UB", "10-UB-PMA", "10-AC", "10-AC-PMA", "10-PC", "10-PC-PMA", "10-FA", "10-FA-PMA", "10-Abd", "10-Abd-PMA", "11-FH", "11-FH-PMA", "11-UB", "11-UB-PMA", "11-AC", "11-AC-PMA", "11-PC", "11-PC-PMA", "11-FA", "11-FA-PMA", "11-Abd", "11-Abd-PMA", "12-FH", "12-FH-PMA", "12-UB", "12-UB-PMA", "12-AC", "12-AC-PMA", "12-PC", "12-PC-PMA", "12-FA", "12-FA-PMA", "12-Abd", "12-Abd-PMA")

col_order <- col_order[ col_order%in%c(colnames(df)) ]

df <- df[, col_order]

```

## Exporting dataframes for RPCA in QIIME2

```{r}

write.table( df[, colnames(df) %in% c( "#OTU ID", colnames(df)[str_detect(colnames(df), "-PMA")])], "/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/Final_rpca_pma.tsv", row.names = FALSE, sep = "\t" , quote = FALSE)

write.table( df[, colnames(df) %in% c( "#OTU ID", colnames(df)[!str_detect(colnames(df), "-PMA")])], "/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/Final_rpca_raw.tsv", row.names = FALSE, sep = "\t" , quote = FALSE)

write.table( df, "/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/Final_rpca_all.tsv", row.names = FALSE, sep = "\t" , quote = FALSE)

```

## Keeping only samples that have matching Raw and PMA treated samples

```{r}

# samples <- c("#OTU ID", "1-AC-PMA", "1-AC", "1-FH-PMA", "1-FH", "1-UB-PMA", "1-UB", "2-AC-PMA", "2-AC", "2-FH-PMA", "2-FH", "2-UB-PMA", "2-UB", "3-FH-PMA", "3-FH", "3-PC-PMA", "3-PC", "3-UB-PMA", "3-UB", "4-AC-PMA", "4-AC", "4-FA-PMA", "4-FA", "4-FH-PMA", "4-FH", "4-PC-PMA", "4-PC", "4-UB-PMA", "4-UB", "5-AC-PMA", "5-AC", "5-FA-PMA", "5-FA", "5-PC-PMA", "5-PC", "5-UB-PMA", "5-UB", "6-AC-PMA", "6-AC", "6-FH-PMA", "6-FH", "6-PC-PMA", "6-PC", "6-UB-PMA", "6-UB", "7-FA-PMA", "7-FA", "7-FH-PMA", "7-FH", "7-PC-PMA", "7-PC", "7-UB-PMA", "7-UB", "8-AC-PMA", "8-AC", "8-Abd-PMA",  "8-Abd",  "8-FA-PMA", "8-FA", "8-FH-PMA", "8-FH", "8-PC-PMA", "8-PC", "8-UB-PMA", "8-UB", "9-AC-PMA", "9-AC", "9-FH-PMA", "9-FH", "9-PC-PMA", "9-PC", "10-Abd-PMA", "10-Abd", "10-FA-PMA", "10-FA", "10-FH-PMA", "10-FH", "10-UB-PMA",  "10-UB",  "11-AC-PMA",  "11-AC", "11-FA-PMA", "11-FA",  "11-FH-PMA", "11-FH", "11-PC-PMA",  "11-PC",  "11-UB-PMA",  "11-UB")
# 
# 
# df <- df[, colnames(df) %in% samples ]

```


## Export

```{r}

write.table(df, "/media/deepan/Deepan/LiveDead/igm-storage2.ucsd.edu/220428_A00953_0539_AHMNLKDSX3/202410_revision/swab_blank_removed/skin_SMGC_Zebra60perc_formatted.tsv", row.names = FALSE, sep = "\t")

```









